Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2020-11-09 write-up/covid-writeup.tex
Update covid-writeup.tex

\usepackage[foot]{amsaddr}
% \usepackage{biblatex}
\usepackage{natbib}
%\bibliographystyle{apalike}
\usepackage{multirow}

@@

Coronavirus case-count data refers to the number of positive molecular tests performed on each day.  Figure~\ref{fig:ny-covid} plots (dotted orange line) the number of reported confirmed COVID-19 cases per day in the state of Indiana.  Total number of COVID-19 diagnostic tests performed per day is also reported (Figure~\ref{fig:ny-covid}, dotted blue line). On August 5th, 2020, for example, there were 1075 total daily positive cases and 11,663 total daily tests.  Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University} and the \href{https://bit.ly/2vUHfrK}{New York Times} provide accessible incoming county-level information of confirmed cases and deaths.

@@

Comparing~\eqref{eq:statdecomp2} to~\eqref{eq:statdecomp} shows that weighting impacts the estimation error in three ways.  First, there is a negative impact on the data quantity component; taking the ratio of these quantities yields
$\sqrt{1 + \frac{CV_W^2}{1-f}} \geq 1$.  Hence, if the data quality does not increase (i.e., $| \rho_{\tilde I (X), Y} | = | \rho_{I,Y}|$ ) then weighting increases the error magnitude.  Note this relative error increase depends on the fraction of population sampled $f$, implying that for large samples there is a larger potential increase in the error if the weights do not improve data quality. Third, the impact of measurement-error on data quality is changed when considering a weighted estimand. {\red state that it may switch sign!als}

@@

\begin{lemma}[IPW Variance] \normalfont
\label{lemma:ipw}
The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
are solutions to the following set of estimating equations
$$
\Phi_n (\eta_t) =
\left (
\begin{array}{c}
\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - (1-FP-FN) \cdot (\mu_t+FP)}{\pi_{j,t}} \\
\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
\end{array}
\right ) = {\bf 0}
$$
where $\eta_t = (\mu_t, \pi_t, FP, FN)$. Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ (check)  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
\end{lemma}

@@

\begin{lemma}[DR Variance] \normalfont
\label{lemma:dr}
The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
are solutions to the following set of estimating equations
$$
\Phi_n (\eta_t) =
\left (
\begin{array}{c}
\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - (1-FP-FN) \cdot (\mu_t+FP)}{\pi_{j,t}} \\
\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
\end{array}
\right ) = {\bf 0}
$$
where $\eta_t = (\mu_t, \pi_t, FP, FN)$. Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ (check)  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
\end{lemma}

\section{Applications to binary outcome and the COVID-19 pandemic.}
\label{section:applications}

What are the propensity measures of importance:
\begin{itemize}
	\item Symptom
	\item Demographics
	\item Prior testing (have you been tested before)
	\item Prevalence (overall worry)
\end{itemize}
There is no good data on individual prior testing so we will ignore.

{\bf Combines death, case count with covariate info, symptoms from Fb,  }

\subsection{COVID-19 active infection prevalence in Indiana}

@@

Here we consider estimation of the active infection rate in Indiana.  Table~\ref{tab:indiana} summarizes data from a random sample from April 25--29th as well as a nonrandom sample obtained between May 2--3 in racial/ethnic minority communities. The nonrandom sample is therefore not a random subsample of the overall case-counts; however, this data provides important supplementary covariate information.  Namely, the samples collected included covariate information on (1) prior symptoms, (2) COVID-19 positive individual in household, and (3) prior COVID-19 positive test. To account for the nonrandom sample being from high risk areas, we adjust estimates to match the statewide positivity rate of 11.7\% between May 1--2. These adjusted rates are presented in parentheses in Table~\ref{tab:indiana}.

\begin{table}[th]
\begin{tabular}{c | c | r r | r | r r }
& & \multicolumn{2}{c}{Total Tests (\%)} & US & \multicolumn{2}{c}{Positive Test Rate (\%)}\\
\cline{3-4} \cline{6-7}
& & NonRandom & Random & Census & NonRandom & Random \\ \hline
\multirow{2}{*}{Sex} & Female & 58.2 & 55.0 & 50.7 & 21.7 (11.2) & 1.4 \\
& Male & 41.8 & 45.0 & 49.3 & 24.2 (12.4) & 2.1 \\ \hline
\multirow{3}{*}{Age} & $<40$ & 39.4 & 28.0 & 52.7 & 29.7 & 1.7 \\
& $40-59$ & 41.1 & 36.0 & 25.2 & 24.9 & 2.1 \\
& $\geq 60$ & 19.5 & 36.0 & 22 & 6.7 & 0.9 \\ \hline
\multirow{2}{*}{Race} & White & 23.1 & 92.0 & 86.9 & 19.5 & 1.5 \\
& Nonwhite & 76.9 & 8.0 & 13.1 & 25.0 & 3.4 \\ \hline
Hispanic & Hispanic & 44.1 & 2.0 & 7.1 & 37.6 & 6.9 \\
Origin & Non-Hispanic & 55.9 & 98.0 & 92.9 & 13.0 & 1.3 \\ \hline
\multirow{2}{*}{Fever} & Yes & 17.0 & 1.8 & - & 66.4 & 4.5 \\
& No & 83.0 & 98.2 & - & 15.6 & 1.3 \\ \hline
Household & Yes & 10.8 & 1.4 & - & 46.1 & 24.4 \\
$+$ Case & No & 89.2 & 98.6 & - & 21.6 & 1.3 \\ \hline
Prior $+$ & Yes & 6.1 & 1.6 & - & 39.2 & 24.4 \\
Test & No & 93.9 & 98.6 & - & 21.6 & 1.3 \\ \hline
\end{tabular}
\caption{Estimated point prevalence of active infection with SARS-CoV-2 by demographics in Indiana.}
\label{tab:indiana}
\end{table}

Comparison to April 20th-25th using Fb data
\begin{itemize}
\item Fever: {\bf FB} = [0.017, 0.020, 0.024], {\bf Random}: 0.018
\item Cough: {\bf FB} = [0.142, 0.148, 0.154], {\bf Random}: 0.149
\item Shortness: {\bf FB} = [0.058, 0.060, 0.062], {\bf Random}: 0.062
\end{itemize}

\subsubsection{Time-varying doubly-robust prevalence estimation}

@@

@@

\subsection{Estimation of Data Quality}
\label{section:est_dq}

\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../methods/figs/NY_casecount.png}
  \caption{NY new cases per day}
  \label{fig:ny-covid-test}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../methods/figs/NY_fracpos.png}
  \caption{NY fraction of new tests positive per day}
  \label{fig:ny-covid-frac}
\end{subfigure}
\caption{New York COVID-19 reported data}
\label{fig:ny-covid}
\end{figure}

>>>>

\usepackage[foot]{amsaddr}
% \usepackage{biblatex}
\usepackage{natbib}
%\bibliographystyle{apalike}
\usepackage{multirow}
\usepackage{enumitem}

@@

Coronavirus case-count data refers to the number of positive molecular tests performed on each day.  Figure~XX plots (dotted orange line) the number of reported confirmed COVID-19 cases per day in the state of Indiana.  Total number of COVID-19 diagnostic tests performed per day is also reported (Figure~XX, dotted blue line). On August 5th, 2020, for example, there were 1075 total daily positive cases and 11,663 total daily tests.  Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University} and the \href{https://bit.ly/2vUHfrK}{New York Times} provide accessible incoming county-level information of confirmed cases and deaths.

@@

Comparing~\eqref{eq:statdecomp2} to~\eqref{eq:statdecomp} shows that weighting impacts the estimation error in three ways.  First, there is a negative impact on the data quantity component; taking the ratio of these quantities yields
$\sqrt{1 + \frac{CV_W^2}{1-f}} \geq 1$.  Hence, if the data quality does not increase (i.e., $| \rho_{\tilde I (X), Y} | = | \rho_{I,Y}|$ ) then weighting increases the error magnitude.  Note this relative error increase depends on the fraction of population sampled $f$, implying that for large samples there is a larger potential increase in the error if the weights do not improve data quality. Third, the impact of measurement-error on data quality is changed when considering a weighted estimand. In particular, weighting may result in $\text{sgn}(\Delta) \neq \text{sgn} (\tilde \Delta)$ which implies the impact of measurement-error may be in a different direction.

@@

\begin{lemma}[Variance of IPW estimator] \normalfont
\label{lemma:ipw}
\textcolor{red}{Must check.}
The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
are solutions to the following set of estimating equations
$$
\Phi_n (\eta_t) =
\left (
\begin{array}{c}
\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - (1-FP-FN) \cdot (\mu_t+FP)}{\pi_{j,t}} \\
\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
\end{array}
\right ) = {\bf 0}
$$
where $\eta_t = (\mu_t, \pi_t, FP, FN)$. Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ (check)  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
\end{lemma}

@@

\begin{lemma}[Variance of doubly-robust estimator] \normalfont
\label{lemma:dr}
\textcolor{red}{Must fix.}
The estimates~$\mu_t := \bar y_{n,t}^\star$ and $\hat \pi_{j,t}$
are solutions to the following set of estimating equations
$$
\Phi_n (\eta_t) =
\left (
\begin{array}{c}
\frac{1}{N} \sum_{j=1}^N I_{j,t} \frac{Y_{j,t} - (1-FP-FN) \cdot (\mu_t+FP)}{\pi_{j,t}} \\
\frac{1}{N} \sum_{t^\prime = 1}^T K_{t,t^\prime} \sum_{j=1}^N I_{j,t^\prime} X_{j,t^\prime} - \frac{1}{N} \sum_{j = 1}^N \tilde I_{j,t^\prime} \tilde W_{j,t^\prime}  \pi_{j,t^\prime} X_{j,t^\prime}  \\
\frac{1}{n_{FP}} \sum_{j \in S_{FP}} \frac{Z_j}{FP} - \frac{1-Z_j}{1-FP} \\
\frac{1}{n_{FN}} \sum_{j \in S_{FN}} \frac{\tilde Z_j}{FN} - \frac{1-\tilde Z_j}{1-FN} \\
\end{array}
\right ) = {\bf 0}
$$
where $\eta_t = (\mu_t, \pi_t, FP, FN)$. Under regularity assumptions (see Appendix~\ref{app:asympderivations}), we have $\bar y_{n,t}^\star - \bar Y_{t} = O_p (n^{-1/2})$ (check)  and $\text{var} (\hat y_{n,t}) = V_{t}^{(IPW)} + o (n^{-1})$ where $V_t^{(IPW)}$ is the first diagonal element of $E [\phi_n(\eta_0)]^{-1} \text{Var}(\Phi_n(\eta_0))E [\phi_n(\eta_0)]^{-1}$.
\end{lemma}

\section{COVID-19 active infection prevalence in Indiana}
\label{section:applications}

@@

Here we consider estimation of the active infection rate in Indiana. We consider five sources of data:
\begin{itemize}[leftmargin=*]
\item {\bf Testing data}:  Each day, the number of tests performed and number of positive tests are reported, with counts broken out jointly by age, gender, and racial demographic information.  Figure~\ref{fig:in-cases} plots the number of COVID-19 positive cases based on testing date by age.

\begin{figure}
\centering
\begin{subfigure}{.45\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../methods/figs/indianacasecounts.png}
  \caption{Indiana case-counts per day}
  \label{fig:in-cases}
\end{subfigure}
\begin{subfigure}{.45\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{../methods/figs/indianadeaths.png}
  \caption{Indiana death-counts per day}
  \label{fig:in-deaths}
\end{subfigure}
\end{figure}

\item {\bf Random/Nonrandom statewide sample}: Table~\ref{tab:indiana} summarizes data from a random sample from April 25--29th as well as a nonrandom sample obtained between May 2--3 in racial/ethnic minority communities (cite). The nonrandom sample is not a random subsample of the overall case-counts; however, this data provides important supplementary covariate information.  Namely, the samples collected included covariate information on (1) prior symptoms, (2) COVID-19 positive individual in household, and (3) prior COVID-19 positive test. To account for the nonrandom sample being from high risk areas, we adjust estimates to match the statewide positivity rate of 11.7\% between May 1--2. These adjusted rates are presented in parentheses in Table~\ref{tab:indiana}.

\begin{table}[th]
\begin{tabular}{c | c | r r | r | r r }
& & \multicolumn{2}{c}{Total Tests (\%)} & US & \multicolumn{2}{c}{Positive Test Rate (\%)}\\
\cline{3-4} \cline{6-7}
& & NonRandom & Random & Census & NonRandom & Random \\ \hline
\multirow{2}{*}{Sex} & Female & 58.2 & 55.0 & 50.7 & 21.7 (11.2) & 1.4 \\
& Male & 41.8 & 45.0 & 49.3 & 24.2 (12.4) & 2.1 \\ \hline
\multirow{3}{*}{Age} & $<40$ & 39.4 & 28.0 & 52.7 & 29.7 (15.0) & 1.7 \\
& $40-59$ & 41.1 & 36.0 & 25.2 & 24.9 (12.5) & 2.1 \\
& $\geq 60$ & 19.5 & 36.0 & 22 & 6.7 (3.4) & 0.9 \\ \hline
\multirow{2}{*}{Race} & White & 23.1 & 92.0 & 86.9 & 19.5 (9.6) & 1.5 \\
& Nonwhite & 76.9 & 8.0 & 13.1 & 25.0 (12.3) & 3.4 \\ \hline
% Hispanic & Hispanic & 44.1 & 2.0 & 7.1 & 37.6 & 6.9 \\
% Origin & Non-Hispanic & 55.9 & 98.0 & 92.9 & 13.0 & 1.3 \\ \hline
\multirow{2}{*}{Fever} & Yes & 17.0 & 1.8 & - & 66.4 (32.1) & 4.5 \\
& No & 83.0 & 98.2 & - & 15.6 (7.5) & 1.3 \\ \hline
Household & Yes & 10.8 & 1.4 & - & 46.1 & 29.4 \\
$+$ Case & No & 89.2 & 98.6 & - & 21.6 & 1.3 \\ \hline
Prior $+$ & Yes & 6.1 & 1.4 & - & 39.2 & 24.4 \\
Test & No & 93.9 & 98.6 & - & 21.6 & 1.3 \\ \hline
\end{tabular}
\caption{Estimated point prevalence of active infection with SARS-CoV-2 by demographics in Indiana.}
\label{tab:indiana}
\end{table}

\item {\bf Facebook symptom survey}: We extract daily symptom surveys completed by individuals on Facebook who identify as living in Indiana.  We collect age, sex, and demographic information as well as COVID-19 related symptoms, which Facebook defines as having (1) a fever and cough, (2) shortness of breath, \emph{or} (3) difficulty breathing. To assess potential bias, we compare the random sample from Indiana to the Facebook symptom survey data collected from April 20th to April 25th.  Table~\ref{tab:comparison} demonstrates minimal bias in the Facebook symptom survey.
\begin{table}
\centering
\begin{tabular}{c | c c c | c}
& \multicolumn{3}{c}{Facebook} & Random \\ \cline{2-5}
& Lower CI & Estimate & Upper CI & Estimate \\ \hline
Fever &  0.017 & 0.020 & 0.024 & 0.018 \\
Cough &  0.142 & 0.148 & 0.154 & 0.149 \\
Shortness & 0.058 & 0.060 & 0.062 & 0.062 \\ \hline
\end{tabular}
\caption{Comparison of Facebook symptom survey and random sample demonstrates minimal bias.}
\label{tab:comparison}
\end{table}

% \item {\bf Hospitalization data}: On each day, we observe the number of individuals who are COVID-19 positive that visited a hospital on that day.  Unfortunately, hospitalization data is not broken out by age, gender, and racial demographic information.  Instead, we rely on combining Facebook data to calculate the fraction of those who sought the hospital per week in each strata. We then use this to figure out how many individuals per strata were hospitalized and then back out how many

\item {\bf Death data}: On each day, we observe the number of deaths due to COVID-19 reported on that day, with deaths broken out jointly by age, gender, and racial demographic information.  Figure~\ref{fig:in-deaths} plots the cumulative number of COVID-19 related deaths over time by age group.
\end{itemize}

\subsection{Inverse-probability weighting approach}

We start with using Table~\ref{tab:indiana} to compute IPW weights.  The random sample is $n=3,658$.  Here, we demonstrate how to compute the weight for a particular strata.  For example, the strata defined as Male, 40--59, White, has a Fever, no household cases, and no prior tests has a weight proportional to
$$
\frac{\left( 0.493 \times 0.252 \times 0.869 \times 0.018 \right)}{\left(0.418 \times 0.411 \times 0.231 \times 0.17 \right)} = 0.628
$$
Using these weights and the adjusted strata-level prevalence estimates, the IPW estimate is $7.7$.

\subsubsection{Time-varying IPW estimator}

\subsection{Model-based estimation}

\subsection{Doubly robust estimation}

@@

% \begin{figure}
% \centering
% \begin{subfigure}{.5\textwidth}
%   \centering
%   \includegraphics[width=.9\linewidth]{../methods/figs/NY_casecount.png}
%   \caption{NY new cases per day}
%   \label{fig:ny-covid-test}
% \end{subfigure}%
% \begin{subfigure}{.5\textwidth}
%   \centering
%   \includegraphics[width=.9\linewidth]{../methods/figs/NY_fracpos.png}
%   \caption{NY fraction of new tests positive per day}
%   \label{fig:ny-covid-frac}
% \end{subfigure}
% \caption{New York COVID-19 reported data}
% \label{fig:ny-covid}
% \end{figure}

@@
